---
title: "Catcam"
author: "Lusia"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    extra_dependencies: ["setspace"]
header-includes: 
  - \usepackage{setspace} 
  - \linespread{2.0}
  - \usepackage{mdframed}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, urlcolor=blue}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(MASS)
library(car)
library(fitdistrplus)
library(goftest)
library(actuar)
library(flexsurv) 
library(truncnorm)
library(pscl)
library(pracma)
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, width.cutoff = 60)
```

# AADT Modeling
The data is obtained from each city (open data, you can click to access the data):

- [Guelph Road](https://explore.guelph.ca/datasets/13eb15489455495b89f65437b05d4f50_1/explore?location=43.452714%2C-80.474796%2C17.70)
- [Kitchener Road](https://open-kitchenergis.opendata.arcgis.com/datasets/c2a28a64955943ffad68e7ce099c5556_0/explore?location=43.458502%2C-80.559454%2C18.70)
- [Bruce County](https://maps.brucecounty.on.ca/arcgisportal/apps/mapviewer/index.html?layers=c9d9b6c118194fff990e749ef0f666da)
- [Ontario](https://www.arcgis.com/apps/mapviewer/index.html?webmap=3402c1b3bcfe485b95f0dec23a1492a2)
- [Waterloo Road](https://data.waterloo.ca/datasets/RMW::-roads/explore?location=43.457658%2C-80.559946%2C18.70)
- [Cambridge Road](https://geohub.cambridge.ca/datasets/21d5954af80e4fa4b45f53ecf201efc3_14/explore?location=43.402358%2C-80.329650%2C12.29)

```{r aadt}
##Name of column and data were modified to ensure consistent formatting (ST=Street)
Kitchener_road <- read_excel("C:/Users/95joo/OneDrive/School-OD/IBIO4521.2/DATA/Kitchener_road.xlsx")
##Removing data causing linearity issue which is also unrelated to our study
Kitchener_road <-subset(Kitchener_road,SPEED<90)
Kitchener_road$OPSCLASS<-as.factor(Kitchener_road$OPSCLASS)
Kitchener_road$LANES<-as.factor(Kitchener_road$LANES)
Kitchener_road$SPEED<-as.factor(Kitchener_road$SPEED)
```
Polynomial model is created using the common parameters found in datasets.

```{r model, echo=TRUE}
Kit_AADT_2<- lm(log(Aadt)~OPSCLASS+DEAD_END+ROADCLASS+LANES+TRUCKROUTE+STREET_TYPE+SPEED, data=Kitchener_road)
vif(Kit_AADT_2)
anova(Kit_AADT_2)
```
Summary of the model:
\begin{verbatim}
Residual standard error: 0.4482 on 5639 degrees of freedom
  (168 observations deleted due to missingness)
Multiple R-squared:  0.9176,	Adjusted R-squared:  0.917 
F-statistic:  1427 on 44 and 5639 DF,  p-value: < 2.2e-16
\end{verbatim}
```{r apply}
Catcam_data <- read_excel("C:/Users/95joo/OneDrive/School-OD/IBIO4521.2/DATA/Catcam_data.xlsx")
Catcam_data <- Catcam_data %>% select_at(vars(1,5,6,9,10))
##Removing "NA" data in duration and AADT
Catcam_filtered <- subset(Catcam_data, Duration !="NA")
Catcam_filtered <- subset(Catcam_filtered, AADT !="NA")
##Changing data format (rounding duration in s, traffic in s)
Catcam_filtered$Duration <- as.numeric(as.character(Catcam_filtered$Duration))
Catcam_filtered$AADT <- as.numeric(as.character(Catcam_filtered$AADT))
Catcam_filtered$time <-round(Catcam_filtered$Duration,digits=0)
Catcam_filtered$traffic <- Catcam_filtered$AADT/(24*60*60)
```

# Variable estimators
A positive integral value dataset, such as frequency and duration, is assumed to follow gamma distribution. Each variables are tested for different distributions, such as log-normal, zero-inflated, and negative binomial. To increase quality of the estimator, multiple distributions of parameters were tested and all data was used instead of cutting out the data. Gamma distribution is estimated using the sample data and it is shown red. Blue graph is sample data distribution. Blue vertical line indicates sample mean and red indicates the expected value.\

## Traffic volume
\begin{verbatim}
  estimate Std. Error
  shape 5.094035 0.3802488
  rate 490.092739 38.4476214
  Loglikelihood: 1358.259 AIC: -2712.518 BIC: -2704.877
  Correlation matrix:
shape rate
shape 1.000000 0.951499
rate 0.951499 1.000000
\end{verbatim}

```{r t_plot, fig.width=4.9, fig.height=3.5}
fit_gamma_t <- fitdist(Catcam_filtered$traffic, "gamma")
simul_gamma_t <- rgamma(10000, shape = fit_gamma_t$estimate["shape"], 
                   rate = fit_gamma_t$estimate["rate"])
ggplot() +
  geom_density(data = data.frame(Value = Catcam_filtered$traffic), aes(x = Value), fill = "blue", alpha = 0.5) +
  geom_density(data = data.frame(Value = simul_gamma_t), aes(x = Value), fill = "red", alpha = 0.5) +
  labs(title = "Gamma Fit vs. Sample data", x = "Traffic (vehicle/s)", y = "Density") +
  geom_vline(aes(xintercept = mean(Catcam_filtered$traffic)), col = "blue")+
  geom_vline(aes(xintercept = mean(simul_gamma_t)), col = "red")
```

## Frequency of crossing
```{r}
Subject <- read_excel("C:/Users/95joo/OneDrive/School-OD/IBIO4521.2/DATA/Catcam_data.xlsx", sheet="Subject")
Subject <- Subject %>% select_at(vars(1:4))
Subject$fre_day<- Subject$frequency/Subject$calendar_day
Subject$fre_week <-Subject$fre_day *7
fit_gamma_f <- fitdist(Subject$fre_week, "gamma")
simul_gamma_f <- rgamma(10000, 
                         shape = fit_gamma_f$estimate["shape"], 
                         rate = fit_gamma_f$estimate["rate"])
```
\begin{verbatim}
  estimate Std. Error
  shape 0.8860221 0.19923256
  rate 0.1048266 0.03112522
  Loglikelihood: -93.88528 AIC: 191.7706 BIC: 195.513
  Correlation matrix:
  shape rate
  shape 1.0000000 0.7571975
  rate 0.7571975 1.0000000
  Min. 1st Qu. Median Mean 3rd Qu. Max.
  0.00004 2.19988 5.68819 8.54521 11.71171 115.11790
\end{verbatim}
```{r f_plot, fig.width=4.9, fig.height=3.5}
ggplot() +
  geom_density(data = data.frame(Value = Subject$fre_week), aes(x = Value), fill = "blue", alpha = 0.5) +
  geom_density(data = data.frame(Value = simul_gamma_f), aes(x = Value), fill = "red", alpha = 0.5) +
  labs(title = "Gamma Fit vs. Sample Data", x = "Frequency of crossing/week", y = "Density") +
  geom_vline(aes(xintercept = mean(Subject$fre_week)), col = "blue")+
  geom_vline(aes(xintercept = mean(simul_gamma_f)), col = "red")
```

# Mortality calculation
Probability of getting hit by a car is multiplied by 0.7, which is the mortality per accident found in epidemiology study and survey. Expected values of each parameter distribution were used to calculate the expected mortality.

## Mortality using v=2m/s, road width=10m
```{r, include=FALSE}
gamma_f<- function(x) {exp(-x)*dgamma(x, shape = fit_gamma_f$estimate["shape"], 
         rate = fit_gamma_f$estimate["rate"])*x}
expected_f <- integral(gamma_f, 0, Inf)
gamma_t<- function(x) {exp(-x)*dgamma(x, shape = fit_gamma_t$estimate["shape"], 
                                      rate = fit_gamma_t$estimate["rate"])*x}
expected_t <- integral(gamma_t, 0, Inf)
# Generate random values from Gamma distributions
F_samples <- rgamma(10000, shape = fit_gamma_f$estimate["shape"], rate = fit_gamma_f$estimate["rate"])
T_samples <- rgamma(10000, shape = fit_gamma_t$estimate["shape"], rate = fit_gamma_t$estimate["rate"])

Mortality <- (1 - exp(-(T_samples) * 0.769)) * 0.7
Mortality_w <-(1 - (exp(-(T_samples) * 0.769)^(F_samples))) * 0.7
fit_gamma_m <- fitdist(Mortality, "gamma")
simul_gamma_m <- rgamma(1000, 
                        shape = fit_gamma_m$estimate["shape"], 
                        rate = fit_gamma_m$estimate["rate"])
```
The expected value of crossing frequency per week and traffic volume per second were used to calculate the weekly mortality of crossing.\
```{r, echo=TRUE}
Success_p<-exp(-expected_t*5)
print(Success_p)
#Per crossing
(1-Success_p)*0.7
#Per week
(1-((Success_p*0.7)^expected_f))
```
## Mortality using v=13m/s, road width=10m
```{r, echo=TRUE}
Success_p2<-exp(-expected_t*0.769)
Mortality_p2 <-1-Success_p2*0.7
print(Success_p2)
print(Mortality_p2)
#Per crossing
(1-Success_p2)*0.7
#Per week
(1-((Success_p2*0.7)^expected_f))
```
## Mortality simulation
Distribution of the mortality per crossing, which follows gamma distribution. The blue vertical line indicates the mean mortality value. \
```{r mortality, fig.width=4.9, fig.height=3.5}
ggplot() +
  geom_density(data = data.frame(Value = simul_gamma_m), aes(x = Value), fill = "red", alpha = 0.5) +
  labs(title = "Gamma Fit of Mortality per Crossing", x = "Mortality per crossing", y = "Density") +
  geom_vline(aes(xintercept = mean(simul_gamma_m)), col = "red")+
  geom_vline(xintercept = 0.0085, col = "blue", show.legend=FALSE)
```
\
Annual mortality was calculated, frequency/year, under assumption of consistent frequency/week throughout the year. 
```{r, echo=TRUE}
##Mortality per crossing summary
summary(Mortality)
##Annual mortality using v=2m/s, road width=10m
(1-((Success_p*0.7)^(expected_f*26)))
##Annual mortality using v=13m/s, road width=10m
(1-((Success_p2*0.7)^(expected_f*26)))
```


