---
title: "GAM workshop"
author: "Lusia"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mgcv) 
library(gratia)
library(kableExtra)
```
Using the built-in CO2 dataset in R, which gives uptake of CO2 in grasses under 
varying CO2 concentrations, with plants from two locations (Mississippi and Quebec)
and two temp treatments; 12 plants, 7 CO2 concentrations for each plant

The default CO2 plant variable is ordered;
This recodes it to an unordered factor (see main text for why)
GAM for non-linear relationship
Spline and basis funtions##
Weight of basis determine spline function##
Choose values for each coefficient to maximize the likelihood
Penaltized likelihood -\lambda to decrease wiggliness
Lambda is based on AIC G... normally REML
Effective degree of freedom is always lower than basis complexity
Therefore, k should be large enough Too large k has high computational cost
Check choice of k with gam.check(), information about residual and p-value etc

```{r, echo=FALSE}
data("CO2")
CO2 <- transform(CO2, Plant_uo=factor(Plant, ordered=FALSE))
```
plot CO2 uptake vs. CO2 concentration for each plant

```{r, echo=FALSE, fig.pos="H"}
(CO2_vis_plot <- ggplot(CO2, aes(x=conc, 
                                 y=uptake, 
                                 group=Plant,
                                 color=Type, 
                                 lty=Plant)) +
   geom_point() +
   geom_line() +
   scale_color_manual(values = rep(c("red","blue","black"), times =4))+
   scale_linetype_manual(values = rep(1:4, each=3))+
   labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})), 
        y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2}))))
```
model log uptake as a function of two smoothers: 
thin plate spline of log concentration and RE for individual-plant-specific intercepts
```{r, echo=FALSE}
CO2_mod1 <- gam(log(uptake) ~ s(log(conc), k=5, bs="tp") +
                  s(Plant_uo, k=12, bs="re"),
                data=CO2, method="REML", family="gaussian")
```
plot the default gratia plot for the CO2 model
```{r, echo=FALSE, fig.pos="H"}
draw(CO2_mod1)
```
left: estimated smoother for effect log CO2 concentration on CO2 uptake
right: QQ plot of the estimated random effects vs. Gaussian quantiles
```{r}
summary(CO2_mod1)
```
Interpreting the summary
edfs represent complexity of the smooth; edf = 1 is a straight line, higher edfs mean more wiggly

Interpretation of significance: a significant smooth is one where you can't 
draw a horizontal line through the 95% CI
high edf =/= significance, and vice versa
a smooth can be linear and significant, non-linear and non-significant, or one of each
```{r}
gam.check(CO2_mod1)
```
makes 4 plots, scroll through them (qq plot, residuals vs. linear predictor, histogram of resids, response vs. fitted)
alternate form of the 4 plots from gratia package:
```{r, echo=FALSE}
appraise(CO2_mod1)
```
just the k part of gam.check
```{r, echo=FALSE}
k.check(CO2_mod1)
```
k for log concentration smooth is probably too low
```{r, echo=FALSE}
table(CO2$conc)
```
but there are only 7 unique values of concentration, so k can be no larger than 7
Setup prediction data (all possible concentration values for all individual plants)
make the prediction, add this and a column of standard errors to the prediction
data.frame. Predictions are on the log scale.
```{r, echo=FALSE}
CO2_mod1_pred <- with(CO2,
                      expand.grid(conc=seq(min(conc), max(conc), length=100),
                                  Plant_uo=levels(Plant_uo)))
CO2_mod1_pred <- cbind(CO2_mod1_pred,
                       predict(CO2_mod1, 
                               CO2_mod1_pred, 
                               se.fit=TRUE))
```
Plot back-transformed model predictions of CO2 uptake by concentration for each plant
```{r, echo=FALSE}
ggplot(data=CO2, aes(x=conc, y=uptake, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(fit - 2*se.fit), ymax=exp(fit + 2*se.fit), x=conc),
              data=CO2_mod1_pred, 
              alpha=0.3, 
              inherit.aes=FALSE) +
  geom_line(aes(y=exp(fit)), data=CO2_mod1_pred) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```
different intercepts but same shape for all plants
functional responses among plants are similar, but model doesn't capture all the patterns
e.g. it underestimates for plant Qc2
fit another model adding on individual-plant-level smoothers
allows shape of the uptake vs. concentration curve to vary between plants (but same wiggliness)
```{r, echo=FALSE}
CO2_mod2 <- gam(log(uptake) ~ s(log(conc), k=5, m=2) + # the "global" smooth for uptake vs. conc
                   s(log(conc), Plant_uo, k=5,  bs="fs", m=2), # the individual-level smooths
                 data=CO2, method="REML")
```
```{r}
draw(CO2_mod2)
summary(CO2_mod2)
gam.check(CO2_mod2)
appraise(CO2_mod2)
```
```{r, echo=FALSE}
CO2_mod2_pred <- predict(CO2_mod2, se.fit=TRUE)
CO2 <- transform(CO2, 
                 mod2 = CO2_mod2_pred$fit, 
                 mod2_se = CO2_mod2_pred$se.fit)
ggplot(data=CO2, aes(x=conc, y=uptake, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(mod2-2*mod2_se),
                  ymax=exp(mod2+2*mod2_se)), alpha=0.25) +
  geom_line(aes(y=exp(mod2))) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```
Same as second model, but now individual-level smoothers can have *different* wiggliness
```{r, echo=FALSE}
CO2_mod3 <- gam(log(uptake) ~ s(log(conc), k=5, m=2, bs="tp") +
                   s(log(conc), by = Plant_uo, k=5, m=1, bs="tp") + # m = 1, so penalize squared first derivative
                   s(Plant_uo, bs="re", k=12), # must explicitly include random intercept
                 data=CO2, method="REML")

# view the smooths
draw(CO2_mod3)
draw(CO2_mod3, select = c(1,14,8,2,11,5), scales = "fixed")
```
```{r}
summary(CO2_mod3)
```
```{r, echo=FALSE}
CO2_mod3_pred <- predict(CO2_mod3, se.fit=TRUE)
CO2 <- transform(CO2, 
                 mod3 = CO2_mod3_pred$fit, 
                 mod3_se = CO2_mod3_pred$se.fit)
ggplot(data=CO2, aes(x=conc, y=uptake, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(mod3-2*mod3_se),
                  ymax=exp(mod3+2*mod3_se)), alpha=0.25) +
  geom_line(aes(y=exp(mod3))) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```
Model AIC comparison
```{r, echo=FALSE}
kbl(AIC(CO2_mod1, CO2_mod2, CO2_mod3))
```
